---
title: "Tidyverse Extend Assignment"
author: "Chinedu Onyeka/Peter Phung"
date: "`r Sys.Date()`"
---

# Introduction

There are two vignettes in this report. The first vignette was created by Chinedu Onyeka and contains several tidyverse operations on a dataset related to heart failure. The second vignette is an extension of the original vignette using additional code written by Peter Phung.


## 1. Tidyverse CREATE from Chinedu Onyeka

## 1A. Vignette Info  
This vignette demonstrates how to use some functions from the dplyr and ggplot2 packages in the tidyverse suit of packages <br/>  

## 1B. Data  
The data used to demonstrate the usage of the dplyr and ggplot2 package was obtained from kaggle. See the link below: <br/>
[Heart Failure Dataset](https://www.kaggle.com/fedesoriano/heart-failure-prediction)

## 1C. Load Libraries  

```{r message=FALSE}
library(tidyverse)
```

## 1D. Read the data  
```{r read-data, message=FALSE}
url <- "https://raw.githubusercontent.com/chinedu2301/DATA607-Data-Acquisition-and-Management/main/heart.csv"
heart_failure <- read_csv(url)
```
```{r head}
# Look into the data to know what columns are present
head(heart_failure)
```
## 1E. dplyr package

### *glimpse*  
Q: What are the variables and data types of the variables in my dataset?  
A: Use dplyr's glimpse function.
```{r}
glimpse(heart_failure)
```

### *select*  
Q: I do not need all the columns, how do I select the columns that I only need for my analysis?  
A: Use dplyr's select function.  
```{r}
# select all columns except Oldpeak and ST_Slope
heart_failure_sub <- heart_failure %>% select(-c(Oldpeak, ST_Slope))
# select columns Age, Sex, RestingBP, and HeartDisease
heart_failure_sub1 <- heart_failure %>% select(Age, Sex, RestingBP, HeartDisease)
# select from columns Age to ExerciseAngina
heart_failure_sub2 <- heart_failure %>% select(Age:ExerciseAngina)
```

View the head of heart_failure_sub
```{r}
head(heart_failure_sub)
```
View the head of heart_failure_sub1
```{r}
head(heart_failure_sub1)
```
View the head of heart_failure_sub2
```{r}
head(heart_failure_sub2)
```
### *filter*  
Q: I want only observations where Sex is M  
A: Use dplyr's filter function.
```{r}
heart_failure_male <- heart_failure %>% filter(Sex == "M")
head(heart_failure_male)
```
Q: I want only observations where RestingECG is Normal  
A: Use dplyr's filter function.  
```{r}
heart_failure_disease <- heart_failure %>% filter(RestingECG == "Normal")
head(heart_failure_disease)
```
## *rename*  
Q: How do I rename a column?  
A: Use dplyr's rename function
```{r}
heart_failure <- heart_failure %>% rename("Gender" = Sex)
head(heart_failure)
```

## 1F. ggplot2 package  

### *Scatterplot*  
Q: I want to plot a scatter plot of RestingBP vs. Cholesterol  
A: Use the ggplot function and the geom_point layer  
```{r}
ggplot(data = heart_failure, aes(x = Cholesterol, y = RestingBP)) + geom_point()
```

Add title, transparency, and theme
```{r}
ggplot(data = heart_failure, aes(x = Cholesterol, y = RestingBP)) + geom_point(alpha = 0.5) + 
  labs(title = "RestingBP vs. Cholesterol") + theme_bw()
```

### *Barplot*  
Q: I want to plot a barchart of the count of Male and Female in the dataset  
A: Use the ggplot function and the geom_bar layer
```{r}
ggplot(data = heart_failure, aes(x = Gender)) + geom_bar(fill = "brown") + 
  labs(title = "Bar chart for count of Sex") + theme_bw()
```

Q: How do I flip co-ordinates?  
A: Add the coord_flip layer
```{r}
ggplot(data = heart_failure, aes(x = Gender)) + geom_bar(fill = "brown") + 
  labs(title = "Bar chart for count of Sex") + theme_bw() + coord_flip()
```


## 2. Tidyverse EXTEND from Peter Phung

Q: I want to colored datapoints by gender for the RestingBP vs. Cholesterol Chart

A: Use the ggplot function again and set the color in the `mapping` to `Gender`.

```{r bolor by gender}

ggplot(data = heart_failure, aes(x = Cholesterol, y = RestingBP, color = Gender)) +
  geom_point()

```

Q: I want different plots of RestingBP vs. Cholesterol for every unique ChestPainType

A: Use `facet_wrap` in order to construct plot separating data by `ChestPainType`

```{r facet_wrap}
ggplot(data = heart_failure, aes(x = Cholesterol, y = RestingBP, color = Gender)) +
  geom_point() +
  facet_wrap(~ChestPainType)
```

Q: Display the counts of Cholesterol for each unique RestingBP with lines instead of bars.

A: Use geom_freqpoly(), which is easier to understand than overlapping multiple historgrams on top of one another.

```{r using geom_freqpoly}

ggplot(data = heart_failure, mapping = aes(x = RestingBP, color = RestingECG)) +
  geom_freqpoly(binwidth = 5)

```

Q: Plot the average cholesterol by age in a line plot

A: Use the `summarize` function to get the mean for each age, then use ggplot to generate the line plot.

```{r summarise}
heart_failure %>%
  filter(ChestPainType == "ATA") %>%
  group_by(Age) %>%
  summarise(mean_cholesterol = mean(Cholesterol, na.rm = TRUE)) %>%
  ungroup() %>%
  ggplot(mapping = aes(x = Age, y = mean_cholesterol)) +
  geom_line()
```

Q: Create columns for each unique ChestPainType using the RestingBP as values for those columns.

A: Use `pivot_wider` to do this

```{r pivot_wider}

pivoted_heart_failure <- heart_failure %>%
  select(ChestPainType, RestingBP) %>%
  group_by(ChestPainType) %>%
  mutate(row = row_number()) %>%
  pivot_wider(names_from = c(ChestPainType), values_from = c(RestingBP)) %>%
  select(-row)

head(pivoted_heart_failure)

```